---
title: "Cross-Validation for Model Selection"
author: "Jessica Kraker"
date: "11/25/2023"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Libraries
```{r, warning=F, message=F}
library(readr)
library(dplyr)
library(ggformula)
library(FNN)
```

### Read in Data
```{r, warning=F, message=F}
Exercise = read_csv("BP_and_exercise.csv")
dim(Exercise)
```

\newpage

# Data Cleaning and Organization - REVIEW

Clean data entry errors
```{r}
Exercise = Exercise %>%
  filter(systolic > diastolic)  # removes 5 points
dim(Exercise)
```

Review and remove impossible data
```{r}
out_diastolic <- Exercise %>%
  filter(diastolic < 10)   # identifies 2 points
out_diastolic

out_sitbend <- Exercise %>%
  filter(`sit and bend forward_cm` > height_cm) # identifies 2 more points

Exercise_no_outliers <- Exercise %>%
  filter(diastolic > 10) %>%
  filter(`sit and bend forward_cm` < height_cm)
dim(Exercise_no_outliers)

fullfit <- lm(systolic ~ .-height_ft -diastolic, 
          data = Exercise_no_outliers)
```

\newpage

# Assessing our model - REVIEW
```{r}
RNGkind(sample.kind = "Rejection")
set.seed(1)

n = nrow(Exercise)
num_train = round(n*2/3)
num_test = n - num_train

groups = c(rep(1, num_train), rep(2, num_test))
random_groups = sample(groups, n)

in_train = (random_groups == 1)

in_test = (random_groups == 2)
```

```{r}
fit_train = lm(systolic ~ .-height_ft -diastolic, 
          data = Exercise[in_train, ])
predictions = predict(fit_train, 
                      newdata = Exercise[in_test, ])
```

MAE
```{r}
mean(abs(Exercise$systolic[in_test] - predictions))
```

\newpage

# NEW - Cross-validation assessment of `fullModel`

Identify and store response variable values:
```{r}
y = Exercise$systolic
```

## Reviewing / Defining the original model
Lesson 1 introduced a `fullModel`: `systolic` related to all other variables except `height_ft` and `diastolic`.  

Define / store that model in R:
```{r}
fullModel <- (systolic ~ .-height_ft -diastolic) # does not need parantheses, but visually helpful
```

Fitting the model:
```{r}
fullfit <- lm(fullModel, data = Exercise)
```

## Honestly assessing the model

Set up the splits of the data - LOOCV:
```{r, eval=F}
nfolds=n
cv_groups = (1:n) 
```

Set up the splits of the data - 10-fold CV:
```{r}
nfolds=10
groups = rep(1:nfolds,length=n)  # nfolds is number of folds
set.seed(2)  
cv_groups = sample(groups,n)  # randomization involved (so we set seed to start consistently)
```


CV process, via for loop, to honestly assess the original model:
```{r}
allpredicted = rep(NA,n)   # storage for honest predictions
ii=1
for (ii in 1: nfolds) {    # ii is an easier string to search for index
  in_train = (cv_groups != ii)     # logical vector indicating NOT group ii
  in_test = (cv_groups == ii)     # logical vector indicating group ii

  train_set = Exercise[in_train,]  # all data EXCEPT for group ii
  test_set = Exercise[in_test, ]   # data in group ii
  
  fit_train <- lm(fullModel, data=train_set) # fit to train set
  
  predicted <- predict(fit_train, newdata = test_set)   # predict for test set
  allpredicted[in_test] <- predicted              # store in locations for group ii
}
```

Computation of honest summary measure `MSE_CV`:
```{r}
MSE_CV = mean((y - allpredicted)^2)
MSE_CV
```
Computation of honest summary measure `MAE_CV`:
```{r}
MAE_CV = mean(abs(y - allpredicted))
MAE_CV
```
Plot of response versus honest predictions:
```{r}
results_fullLM_CV <- data.frame(allpredicted, y)
gf_point(y ~ allpredicted, data=results_fullLM_CV) |>
  gf_abline(intercept=0, slope=1, color="orange3", linewidth = 2) |>
  gf_abline(intercept=0-MAE_CV, slope=1, color="orange", linetype = 2) |>
  gf_abline(intercept=0+MAE_CV, slope=1, color="orange", linetype = 2)
```


\newpage

# NEW - compare various linear regression models

Linear Model specifications 
```{r}
Model1 = (systolic ~ .-height_ft -diastolic) 
Model2 = (systolic ~ age + `broad jump_cm` + height_ft + gender)
Model3 = (systolic ~ age + `broad jump_cm` + gender + age:gender + `broad jump_cm`:gender)
Model4 = (systolic ~ age + `broad jump_cm` + gender + class + age:gender + `broad jump_cm`:gender + age:class + `broad jump_cm`:class)

allModels = list(Model1,Model2,Model3,Model4)
allModels[[1]]
```
## CV process, use m as index for model
```{r}
allpredicted = rep(NA,n)   # storage for honest predictions

m = 1
# indexed model in cross-validation
for (ii in 1: nfolds) {    # ii is an easier string to search for index
  in_train = (cv_groups != ii)     # logical vector indicating NOT group ii
  in_test = (cv_groups == ii)     # logical vector indicating group ii

  train_set = Exercise[in_train,]  # all data EXCEPT for group ii
  test_set = Exercise[in_test, ]   # data in group ii
  
  fit_train <- lm(allModels[[m]], data=train_set) # fit to train set
  
  predicted <- predict(fit_train, newdata = test_set)   # predict for test set
  allpredicted[in_test] <- predicted              # store in locations for group ii
}

MAE_CV = mean(abs(y - allpredicted))
MAE_CV

results_LM_m_CV = data.frame(y, allpredicted)  # allpredicted will differ depending on model m
gf_point(y ~ allpredicted, data=results_LM_m_CV,
         title = paste("Compare Observed to Predicted for Model",m),
         ylab = "Observed", xlab = "Predicted")
```

## CV process streamlined, loop through m as index for model
```{r}
# set up storage for results
allMAEvaluesCV = rep(NA,length(allModels))
allMSEvaluesCV = rep(NA,length(allModels))
allplots <- list()

# for-loop to cycle through models
for (m in 1:length(allModels)) {
  allpredicted = rep(NA,n)   # storage for honest predictions
  # revised loop to do cross-validation
  for (ii in 1: nfolds) {    # ii is an easier string to search for index
    in_train = (cv_groups != ii)     # logical vector indicating NOT group ii
    in_test = (cv_groups == ii)     # logical vector indicating group ii

    train_set = Exercise[in_train,]  # all data EXCEPT for group ii
    test_set = Exercise[in_test, ]   # data in group ii
  
    fit_train <- lm(allModels[[m]], data=train_set) # fit to train set
  
    predicted <- predict(fit_train, newdata = test_set)   # predict for test set
    allpredicted[in_test] <- predicted              # store in locations for group ii
  }
  
  allMSEvaluesCV[m] = mean((y - allpredicted)^2)
  allMAEvaluesCV[m] = mean(abs(y - allpredicted))

  results_LM_m_CV = data.frame(y, allpredicted)  # allpredicted will differ depending on model m
  allplots[[m]] <- gf_point(y ~ allpredicted, data=results_LM_m_CV,
         title = paste("Compare Observed to Predicted for Linear Model",m),
         ylab = "Observed", xlab = "Predicted")
}
```

## Review Results - Identify best model
```{r}
allMAEvaluesCV  # Model 1 is "best"
modelnums = (1:4)

gf_point(allMSEvaluesCV ~ modelnums)
gf_point(allMAEvaluesCV ~ modelnums)
```

- Describe best model and visualize fit
```{r}
which.min(allMAEvaluesCV)
minMAEvalueCV_linear <- allMAEvaluesCV[which.min(allMAEvaluesCV)]; minMAEvalueCV_linear
bestlinearmodel <- allModels[[which.min(allMAEvaluesCV)]]; bestlinearmodel
allplots[[which.min(allMAEvaluesCV)]]
```

- fit best model to full data
```{r}
fullfit <- lm(bestlinearmodel, data=Exercise)
coef(fullfit)
```

\newpage

# NEW - compare various kNN regression models

Additional predictor preparation - one-hot encode and remove original
```{r}
str(Exercise)
table(Exercise$gender)
table(Exercise$class)

Exercise <- mutate(Exercise,
                  # genderM: indicate "M", with baseline "F"
                  genderM=ifelse(gender == "M", 1, 0),
                  # classA, classB, classC, classD: indicate each category of class
                  classA=ifelse(class == "A", 1, 0),
                  classB=ifelse(class == "B", 1, 0),
                  classC=ifelse(class == "C", 1, 0),
                  classD=ifelse(class == "D", 1, 0),
                  gender=NULL, class=NULL)
head(Exercise)

# remove other BP measure and original-unit height_ft
Exercise <- select(Exercise, -diastolic, -height_ft)
head(Exercise)
                
# check
names(Exercise)
table(Exercise$genderM)
table(Exercise$classA)
table(Exercise$classB)
table(Exercise$classC)
table(Exercise$classD)
```

kNN Model specifications 
```{r}
# n is quite large
sqrt(n)
# try integers 1-20, then in steps of 10 up to 230 (double the square root)
# kvals <- c(1:20,c(3:23)*10); kvals
kvals <- c(c(1:23)*10); kvals

# could also consider subsets of predictors
```


## CV process streamlined, loop through m as index for model
```{r}
# set up storage for results
allMAEvaluesCV_kNN = rep(NA,length(kvals))
allMSEvaluesCV_kNN = rep(NA,length(kvals))
allplots <- list()

# for-loop to cycle through models
for (m in 1:length(kvals)) {
  allpredicted = rep(NA,n)   # storage for honest predictions
  
  # revised loop to do cross-validation
  for (ii in 1: nfolds) {    # ii is an easier string to search for index
    in_train = (cv_groups != ii)     # logical vector indicating NOT group ii
    in_test = (cv_groups == ii)     # logical vector indicating group ii

    quant_train_std = scale(Exercise[in_train, c(1:2,4:7)]) # quantitative predictors in train set
    quant_test_std = scale(Exercise[in_test, c(1:2,4:7)], 
                           center = attr(quant_train_std, "scaled:center"),
                           scale = attr(quant_train_std, "scaled:scale"))
    x_train = cbind(quant_train_std,
                    Exercise[in_train, c(8:12)])
    x_test = cbind(quant_test_std,
                    Exercise[in_test, c(8:12)])

    knnfit <- knn.reg(x_train, x_test, y=Exercise$systolic[in_train], k=kvals[m])
    predicted <-  knnfit$pred  # predict for test set
    allpredicted[in_test] <- predicted              # store in locations for group ii
  }
  
  allMSEvaluesCV_kNN[m] = mean((y - allpredicted)^2)
  allMAEvaluesCV_kNN[m] = mean(abs(y - allpredicted))

  results_kNN_m_CV = data.frame(y, allpredicted)  # allpredicted will differ depending on model m
  allplots[[m]] <- gf_point(y ~ allpredicted, data=results_kNN_m_CV,
         title = paste("Compare Observed to Predicted for kNN Model",m),
         ylab = "Observed", xlab = "Predicted")
  print(kvals[m])
}
```


## Review Results - Identify best model
```{r}
allMAEvaluesCV_kNN  # Model  with k=150 is "best"

gf_point(allMSEvaluesCV_kNN ~ kvals)
gf_point(allMAEvaluesCV_kNN ~ kvals)
```

## Review Results - Identify best model
```{r}
which.min(allMAEvaluesCV_kNN)
minMAEvalueCV_kNN <- allMAEvaluesCV_kNN[which.min(allMAEvaluesCV_kNN)]; minMAEvalueCV_kNN
bestk <- kvals[which.min(allMAEvaluesCV_kNN)]; bestk
allplots[[which.min(allMAEvaluesCV_kNN)]]  |>
  gf_abline(intercept=0, slope=1, color="orange3", linewidth = 1) |>
  gf_abline(intercept=0-minMAEvalueCV_kNN, slope=1, color="orange", linetype = 2) |>
  gf_abline(intercept=0+minMAEvalueCV_kNN, slope=1, color="orange", linetype = 2)
```

# Finally, compare between methods
```{r}
all_methods <- c("Linear", "kNN")
all_best_hyperparameters <- list(bestlinearmodel, bestk)
all_methods_min_CV <- c(minMAEvalueCV_linear, minMAEvalueCV_kNN)
all_methods[which.min(all_methods_min_CV)]
all_best_hyperparameters[[which.min(all_methods_min_CV)]]
```

